---
title: "Introduction to crossRanger"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to crossRanger}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4,
                      message = FALSE, warning = FALSE)
```

crossRanger builds random forest (RF) models within multiple microbiome datasets, transfers
the models across datasets, and discovers generalizable biomarkers. This vignette walks through
the four modules of the package with a simulated multi-cohort case-control dataset.

```{r}
library(crossRanger)
data(crossRanger_sim)
abundance <- crossRanger_sim$abundance
metadata <- crossRanger_sim$metadata
table(metadata$cohort, metadata$disease_status)
```

Each subject contributes two samples (visits), which are correlated. In all cohorts, cases are
enriched in `g__Genus01-03` and depleted in `g__Genus04-05`; each cohort also has one cohort-specific marker.

## Module 1: within-study modeling

`rf_clf.by_datasets()` trains one classifier per dataset (`s_category`) for the target
(`c_category`). Group k-fold cross-validation keeps both samples of a subject in the same fold,
preventing information leakage.

```{r, results = "hide"}
res <- rf_clf.by_datasets(abundance, metadata, s_category = "cohort", c_category = "disease_status",
                          positive_class = "Case", nfolds = 5, cv_type = "group",
                          g_category = "subject_id", ntree = 200, n_cores = 2)
```

```{r}
res$perf_summ[, c("Dataset", "Sample_size", "Validation_type", "AUROC", "AUPRC", "Accuracy", "F1")]
plot_clf_ROC(res)
```

Feature tables from separate studies can also be supplied as a list; they are harmonized to
the shared features with `harmonize_features()`.

## Module 2: cross-dataset transfer

`rf_clf.cross_appl()` applies each model to every other dataset. `rf_clf.lodo()` trains on all
datasets but one and tests on the held-out dataset.

```{r, results = "hide"}
cross_rf <- rf_clf.cross_appl(res$rf_model_list, res$x_list, res$y_list, positive_class = "Case")
lodo <- rf_clf.lodo(abundance, metadata, s_category = "cohort", c_category = "disease_status",
                    positive_class = "Case", ntree = 200, n_cores = 2)
```

```{r}
plot_cross_appl(cross_rf, metric = "AUROC")
lodo$perf_summ[, c("Test_data", "Train_size", "Test_size", "AUROC", "Accuracy")]
```

## Module 3: feature selection and biomarker discovery

Recursive feature elimination (RFE) identifies the smallest feature set whose cross-validated
AUROC is within 1% of the best one.

```{r, results = "hide"}
cohort_a <- metadata$cohort == "Cohort_A"
x_a <- abundance[cohort_a, ]
y_a <- metadata$disease_status[cohort_a]
rfe <- rf_clf.rfe(x_a, y_a, nfolds = factor(group.folds(metadata$subject_id[cohort_a], 5)),
                  metric = "AUROC", tolerance = 0.01, ntree = 200)
```

```{r}
plot_clf_feature_selection(rfe)$plot
rfe$selected_features
```

Univariate tests on CLR-transformed data complement the ML-derived rankings.
Features both selected by RFE and significant in the tests form a robust biomarker set.

```{r}
bg_test <- BetweenGroup.test(x_a, y_a, clr_transform = TRUE, p.adj.method = "fdr", q_cutoff = 0.05)
markers <- id_robust_markers(rfe, bg_test)
head(markers[, c("feature", "rf_selected", "non.param.test_p.adj", "mean_logfc", "Enr", "robust_marker")])
```

Across datasets, the log2 fold change heatmap and the consistency of importance rankings separate
universal markers from cohort-specific ones.

```{r, fig.height = 5}
plot_logfc_heatmap(res, min_sig_datasets = 1)$plot
consistency <- corr_datasets_by_imps(res, top_n = 5)
head(consistency$feature_consistency)
```

## Module 4: regression, visualization and reports

The regression functions share the same infrastructure, e.g., predicting age within each cohort:

```{r, results = "hide"}
reg <- rf_reg.by_datasets(abundance, metadata, s_category = "cohort", c_category = "age",
                          nfolds = 5, cv_type = "group", g_category = "subject_id",
                          ntree = 200, n_cores = 2)
```

```{r}
reg$perf_summ[, c("Dataset", "MAE", "RMSE", "R_squared", "Spearman_rho")]
```

All results can be combined into a reproducible HTML, PDF or Word report:

```{r, eval = FALSE}
crossRanger_report(Within_study = res, Cross_application = cross_rf, LODO = lodo,
                   RFE_Cohort_A = rfe, Age_regression = reg, outdir = "crossRanger_results")
```
